Improving the Fidelity of Redox Flow Battery Schedule Optimisation for more Accurate Techno-economic Analyses.

### 2018-11-19 Paper/main.tex

\title{The importance of fidelity in the scheduling of Redox Flow Batteries.}

@@

\begin{abstract}
Redox Flow Battery (RFB) systems are promising technologies for the multi-hour electrical energy storage that would be necessary in a future low carbon emission UK electrical grid based on wind and solar power. In order to perform techno-economic analyses (TEA) on these systems, it is necessary to calculate potential revenue for realistic applications, ideally under optimal operation (scheduling). In the present work, a novel schedule optimisation framework is introduced in which coulombic and voltaic components are separated, and the problem variablised in terms of current-density rather than power. This approach allows the expression of ohmic losses as a convex quadratic function of current-density in the objective function while maintaining all linear constraints. The resultant non-linear problem is more readily solvable than previous expressions in terms of power. It is demonstrated for an arbitrage application involving a vanadium RFB that the consideration of variable voltaic losses results in a significant improvement to the accuracy of revenue estimation when compared to the typical linear programming approach, where a fixed efficiency is assumed. The framework allows the removal of the assumption of a representative fixed open-cell voltage, which further improves revenue estimation, at the cost of computability. In a third case study, the framework allows a linear constraint on the working cell-voltage, resulting in current-density restriction at high state of charge (SOC) in order to avoid hydrogen and oxygen evolution. This feature would be relevant for the optimisation of Li-ion battery scheduling, where voltage is a very important consideration. The introduced framework is intended to assist technical optimisation of future RFB systems, for example in determining the optimum pump or cooling system size. The framework allows product optimisation in terms of revenue under optimal scheduling against a price signal (rather than metrics such as system cost, or lifetime cost per unit energy throughput as in previous works).

@@

\begin{keyword}
Redox flow Battery \sep Optimisation \sep Non-linear Programming \sep Techno-Economic Analysis \sep Scheduling \sep VRFB  
%% keywords here, in the form: keyword \sep keyword

@@

%% main text
\section{Introduction}
\label{sec:Intro}
Worldwide, the need to reduce CO$_2$ emissions is stimulatling the development of a wide range of renewable energy sources. Presently, renewable energy is estimated to account for 10\% of global energy consumption, and this is forecast to grow to 12\% by 2023. The growth of renewable energy is forecast to be highest in the electrical power sector, from 24\% in 2017 to 30\% in 2023. Although hydro power presently accounts for 50\% of renewable power generation, the  intermittent power sources wind and solar are forecast to contribute most to this growth\cite{IEA2018}. A number of large European economies, such as Germany, UK, Italy and Spain obtained 23\% to 30\% of there electrical energy from non-hydro renewable power in 2017 \cite{BP2018}.

@@

In order to quantify the economic merits of RFB systems in comparison to other BESS, or indeed ESS, it is necessary to perform techno-economic analyses (TEA) for particular applications. TEA should involve optimisation of the operation of a BESS to maximise revenue (or minimise costs). This should be done at various energy and power ratings (as these are decoupled in a non-hybrid RFB) to find the global optimum, as described by \cite{Oudalov2007}. The optimisation of operation requires determining the schedule, that is, the power output of the battery in each sub-period that maximises benefit without violating energy conservation.

Where a price signal exists optimisation is commonly performed by posing a linear programming (LP) problem \emph{do I dump here a list of all the LP references i will refer to below?}. In this method, the objective function and the constraints on the system are expressed as linear algebraic terms. On the resulting geometric surface, the optimum is guaranteed to lie on a vertex, hence a time-consuming search of the whole surface is not required.

@@

\cite{Hu2010} optimised the schedule of a VRFB and a poly-sulphide bromine RFB in order to maximise arbitrage revenue on the Danish day ahead electrical spot market. This approach was not strictly LP, as discussed in \cref{Model_Formulation_LP_Reference_Current_Density_Terms}, but the problem may be reposed in LP form.

@@

\cite{Sarker2017} also incorporated the dependency of efficiency on power in a scheduling problem (although applied to a Li-ion battery rather than a RFB). Quadratic terms are employed to relate losses to C-rate of charge and discharge. The presence of this quadratic function on the denominator of the discharge SOC increment term is avoided by redefining the discharge power variable as upstream of losses (rather than the power delivered to the grid). The discharge efficiency function is then moved from the SOC expression and multiplied by the discharge power variable in the objective function. The charge efficiency is left in the SOC expression (redefining the charge power variable in the same way would require dividing it by the quadratic charge efficiency term in the objective function). The problem therefore becomes convex (quadratic), but non-linear in both objective function and constraint. The authors then employ a linear approximation technique\sol{does this need a reference}, and combine the efficiency treatment with a piece-wise linear treatment of degradation as a function of power, giving an overall mixed integer-linear programming problem (MILP).

Accurate treatment of efficiency is important, as losses impact the revenue available from arbitrage. The key innovation of the present work is the introduction of a RFB schedule optimisation framework posed in terms of current-density rather than power. This allows the separation of the voltaic and coulombic components of the power input/output. Using the introduced framework, three novel augmentations \emph{(add-ons?)} to the typical scheduling problem are posed while maintaining an easily computable problem:

@@

The importance of these improvements to the fidelity of revenue estimation is demonstrated in a day-ahead price arbitrage case study.

@@

The SOC of the RFB is defined after \cite{Gomes2017} by:

@@

The performance of arbitrage between 16:00 and 20:00 is notably reduced in \cref{fig:OCV}. In this period there are small fluctuations in price from which the RFB derived revenue when a fixed OCV was assumed. However the global optimum requires that the RFB be at a low SOC in these hours, and hence at a low OCV. The low OCV means that a higher current-density would be required to obtain a given pwoer outpute, and hence higher voltaic losses would be incurred, eroding the revenue. On this date the impact on revenue is low; the optimal schedule with consideration of functional OCV results in a revenue of \pounds 0.062 compared to \pounds 0.063 for the base case. Across 2017 the formulation with functional OCV returns \pounds 26.40 compared to \pounds 27.81 for the fixed OCV case. The assumption of a fixed OCV (parametrised at 50\% SOC) therefore results in a small but not insignificant overestimation of revenue in this application. Including functionalised OCV in the objective function increases the solve time - the 2017 data were processed in \SI{217}{\second} compared to \SI{129}{\second} for the convex objective function in \cref{eqn: Non-Linear_Schedule_Objective_Function_Current_as_Variable}.

@@

\cref{fig:V_max_schedules_no_effect_day} shows the optimal scheduling with and without voltage constraint for the median revenue day 22\textsuperscript{nd} August 2017, and \cref{fig:V_max_voltages_no_effect_day} the corresponding cell voltage profiles, showing a maximum cell-voltage of \SI{1.62}{\volt} and hence no violation in the unconstrained schedule.

\begin{figure}[!ht]
\centering
\subfloat[][]{
\label{fig:V_max_schedules_no_effect_day}
\includegraphics[trim = 2cm 8cm 2cm 9cm, clip, width = .5\textwidth]{./figures/V_max_schedules_no_effect_day.pdf}
}
\subfloat[][]{
\label{fig:V_max_voltages_no_effect_day}
\includegraphics[trim = 2cm 8cm 2cm 9cm, clip, width = .5\textwidth]{./figures/V_max_voltages_no_effect_day.pdf}
}
\caption{(a) Optimal VRFB schedules derived for the 22\textsuperscript{nd} August 2017 by maximising \cref{eqn: NLP_true_OCV_long_form} with no cell-voltage constraint, and subject to the cell voltage constraint given in \cref{eqn: Cell_Voltage_Constraint}. (b) Corresponding cell-voltage profiles.}
\end{figure}

\cref{fig:V_max_schedules} shows the optimal schedules with and without constraint for 5\textsuperscript{th} February 2017, where the difference in revenue was greatest (\pounds 0.018 vs. \pounds 0.020). \cref{fig:V_max_voltages} shows the corresponding cell-voltage profiles.

@@

In the unconstrained case the RFB charges at the maximum current-density of \SI{3200}{\ampere\per\square\meter} in hour 19 in order to capitalise on the unusually low electrical price, resulting in a cell-voltage of \SI{1.71}{\volt}. In the constrained case, the current-density is restricted to \SI{2213}{\ampere\per\square\meter} in the same period, and the RFB must charge more (and discharge less) in other periods to satisfy \cref{eqn: Method_Constraint_E_Conservation}. The impact of the constraint would be more severe if the variations in price data were greater, or the maximum voltage set at a lower value.

The application of the linear constraint on cell-voltage increases the time to process the 2017 data from \SI{217}{\second} to \SI{232}{\second}.

An important aspect of RFB economics is the variability of daily price profile across the year. Figure XX shows composite day-ahead price data for the winter and summer seasons in 2017. In summer months, the availability of cheap solar power depresses the price around midday, and the peak evening demand is low, resulting in two peaks of similar height. In winter, the solar price depression is not so marked, and the price in the evening peak becomes higher, hence the profile is more like a single peak.

\cref{fig:winter_day_schedules,fig:summer_day_schedules} show the optimal LP and NLP (\cref{Model_Formulation_Constraint_Cell_Voltage}) schedules for example winter and summer days from 2017. Although the revenue obtained is \pounds 0.074 for both days under the NLP optimal schedule, the required throughput of the battery is greater on the summer day. This has an impact on the error of the LP revenue estimation - for the winter day the revenue is estimated at \pound 0.088, whereas on the summmer day it is estimated at \pound 0.099. The error in the estimation of revenue under the LP optimisation is dependent on energy throughput.

>>>>

\title{Improving the Fidelity of Redox Flow Battery Schedule Optimisation for more Accurate Techno-economic Analyses.}

@@

\begin{abstract}
Redox Flow Battery (RFB) systems are promising technologies for the multi-hour electrical energy storage that would be necessary in a future low carbon emission  electrical grids based on wind and solar power. In order to perform techno-economic analyses (TEA) on these systems it is necessary to calculate potential revenue for realistic applications, ideally under optimal operation (scheduling). In this work a novel schedule optimisation framework is introduced in which coulombic and voltaic components are separated, and the problem variablised in terms of current-density rather than power. This framework allows the expression of ohmic losses as a convex quadratic function of current-density in the objective function while maintaining all linear constraints. It is demonstrated for an arbitrage application involving a vanadium RFB that the consideration of variable voltaic losses results in a significant improvement to the accuracy of revenue estimation when compared to the typical linear programming (LP) approach where a fixed efficiency is assumed. The quadratic formulation is more readily solvable than alternative approaches reported in the literature, with minimal increase to solve time compared to the LP approach. The fidelity of the RFB representation is further increased by expressing open cell voltage as a function of state of charge (SOC). At low SOC, the current-density required for a given power is greater, and hence the ohmic losses are higher. In a third development, a linear constraint is applied to the working cell-voltage, resulting in current-density restriction at high SOC in order to avoid hydrogen and oxygen evolution. This feature would also be useful for schedule optimisation of Li-ion batteries, where voltage is an important consideration. The introduced framework is intended to assist techno-economic optimisation of future RFB systems, for example in determining the optimum pump or cooling system size for a particular application. The framework allows system optimisation in terms of revenue under optimal scheduling against a price signal rather than metrics such as system cost, or lifetime cost per unit energy throughput as in previous works.

@@

\begin{keyword}
Redox Flow Battery \sep Optimisation \sep Non-linear Programming \sep Techno-Economic Analysis \sep Scheduling \sep VRFB  
%% keywords here, in the form: keyword \sep keyword

@@

%% main text
\section{Introduction}
\label{sec:Intro}
Worldwide, the need to reduce CO$_2$ emissions is stimulating the development of a wide range of renewable energy sources. Presently, renewable energy is estimated to account for 10\% of global energy consumption, and this is forecast to grow to 12\% by 2023. The growth of renewable energy is forecast to be highest in the electrical power sector, from 24\% in 2017 to 30\% in 2023. Although hydro power presently accounts for 50\% of renewable power generation, the  intermittent power sources wind and solar are forecast to contribute most to this growth\cite{IEA2018}. A number of large European economies, such as Germany, UK, Italy and Spain obtained 23\% to 30\% of there electrical energy from non-hydro renewable power in 2017 \cite{BP2018}.

@@

In order to quantify the economic merits of RFB systems in comparison to other BESS, or indeed ESS, it is necessary to perform techno-economic analyses (TEA) for particular applications. TEA should involve optimisation of the operation of a BESS to maximise revenue (or minimise costs). This should be done at various energy and power ratings (as these are decoupled in a non-hybrid RFB) to find the global optimum, as described by \cite{Oudalov2007}. The optimisation of operation requires determining the schedule, that is the power output of the battery in each sub-period, that maximises benefit without violating energy conservation.

Where a price signal exists optimisation is commonly performed by posing a linear programming (LP) problem. In this method, the objective function and the constraints on the system are expressed as linear algebraic terms. On the resulting geometric surface, the optimum is guaranteed to lie on a vertex, hence a time-consuming search of the whole surface is not required.

@@

\cite{Hu2010} optimised the schedule of a VRFB and a poly-sulphide bromine RFB in order to maximise arbitrage revenue on the Danish day ahead electrical spot market. This approach was not strictly LP, but the problem may be reposed in LP form.

@@

\cite{Sarker2017} also incorporated the dependency of efficiency on power in a scheduling problem (although applied to a Li-ion battery rather than a RFB). Quadratic terms are employed to relate losses to C-rate of charge and discharge. The presence of this quadratic function on the denominator of the discharge SOC increment term is avoided by redefining the discharge power variable as upstream of losses (rather than the power delivered to the grid). The discharge efficiency function is then moved from the SOC expression and multiplied by the discharge power variable in the objective function. The charge efficiency is left in the SOC expression (redefining the charge power variable in the same way would require dividing it by the quadratic charge efficiency term in the objective function). The problem therefore becomes convex (quadratic), but non-linear in both objective function and constraint. The authors then employ a linear approximation technique\sol{does this need a reference? DR: do you mean for the technique used?}, and combine the efficiency treatment with a piece-wise linear treatment of degradation as a function of power, giving an overall mixed integer-linear programming problem (MILP).

Accurate treatment of efficiency is important, as losses impact the revenue available from arbitrage. The key innovation of the present work is the introduction of a RFB schedule optimisation framework posed in terms of current-density rather than power. This allows the separation of the voltaic and coulombic components of the power input/output. Using the introduced framework, three improvements to the RFB representation in the schedule optimisation are posed while maintaining an easily computable problem:

@@

The importance of these improvements to the accuracy of revenue estimation is demonstrated in a day-ahead price arbitrage case study.

@@

The SOC of the RFB at time $t$ is defined after \cite{Gomes2017} by:

@@

The performance of arbitrage between 16:00 and 20:00 is notably reduced in \cref{fig:OCV}. In this period there are small fluctuations in price from which the RFB derived revenue when a fixed OCV was assumed. However the global optimum requires that the RFB be at a low SOC in these hours, and hence at a low OCV. The low OCV means that a higher current-density would be required to obtain a given pwoer outpute, and hence higher voltaic losses would be incurred, eroding the revenue. On this date the impact on revenue is low; the optimal schedule with consideration of functional OCV results in a revenue of \pounds \SI{0.062}{\per\kilo\watt} compared to \pounds \SI{0.063}{\per\kilo\watt} for the base case. Across 2017 the formulation with functional OCV returns \pounds \SI{26.40}{\per\kilo\watt} compared to \pounds \SI{27.81}{\per\kilo\watt} for the fixed OCV case. The assumption of a fixed OCV (parametrised at 50\% SOC) therefore results in a small but not insignificant overestimation of revenue in this application. Including functionalised OCV in the objective function increases the solve time - the 2017 data were processed in \SI{217}{\second} compared to \SI{129}{\second} for the convex objective function in \cref{eqn: Non-Linear_Schedule_Objective_Function_Current_as_Variable}.

@@

\cref{fig:V_max_schedules} shows the optimal schedules with and without constraint for 5\textsuperscript{th} February 2017, where the difference in revenue was greatest (\pounds \SI{0.18}{\per\kilo\watt} vs. \pounds \SI{0.20}{\per\kilo\watt}). \cref{fig:V_max_voltages} shows the corresponding cell-voltage profiles.

@@

In the unconstrained case the RFB charges at the maximum current-density of \SI{3200}{\ampere\per\square\meter} in hour 20 in order to capitalise on the unusually low electrical price, resulting in a cell-voltage of \SI{1.71}{\volt}. In the constrained case, the current-density is restricted to \SI{2213}{\ampere\per\square\meter} in the same period, and the RFB must charge more (and discharge less) in other periods to satisfy \cref{eqn: Method_Constraint_E_Conservation}. The impact of the constraint would be more severe if the variations in price data were greater, or the maximum voltage set at a lower value.

The application of the linear constraint on cell-voltage increases the time to process the 2017 data from \SI{217}{\second} to \SI{232}{\second}.
\subsection{Impact of price profile on optimisation results}
\label{Results_impact_of_price_profile}
An important aspect of RFB economics is the variability of daily price profile across the year. \Cref{fig:DJF_2017_N2EX,fig:JJA_2017_N2EX} show composite day-ahead price data for the winter and summer seasons in 2017. In summer months, the availability of cheap solar power depresses the price around midday, and the peak evening demand is low, resulting in two peaks of similar height. In winter, the solar price depression is not so marked, and the price in the evening peak becomes higher, hence the profile is more like a single peak.

\begin{figure}[!ht]
\centering
\subfloat[][]{
\label{fig:JJA_2017_N2EX}
\includegraphics[trim = 13.8cm 3.5cm 10.3cm 4.5cm, clip, width = .4\textwidth]{./figures/JJA_2017.pdf}
}
\subfloat[][]{
\label{fig:DJF_2017_N2EX}
\includegraphics[trim = 13.8cm 3.5cm 10.3cm 4.5cm, clip, width = .4\textwidth]{./figures/DJF_2017.pdf}
}
\caption{Composite hourly N2EX day-ahead price data for a) June, July and August 2017 and b) December,January, February 2017}
\end{figure}

\cref{fig:winter_day_schedules,fig:summer_day_schedules} show the optimal LP and NLP (\cref{Model_Formulation_Constraint_Cell_Voltage}) schedules for example winter and summer days from 2017. Although the revenue obtained is \pounds \SI{0.074}{\per\kilo\watt} for both days under the NLP optimal schedule, the throughput of the RFB is greater on the summer day. This has an impact on the error of the LP revenue estimation; for the winter day the revenue is estimated at \pounds \SI{0.088}{\per\kilo\watt}, whereas on the summmer day it is estimated at \pounds \SI{0.099}{\per\kilo\watt}. The error in the estimation of revenue using the LP optimisation is therefore dependent on the application.
